---
title: "Effects of Gratitude Journaling on Well-being"
author: "Ronak Fathi"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(mice)
library(broom)
```

# Introduction
Gratitude journaling has been associated with improved emotional well-being and reduced stress. This study simulates a 2-week longitudinal intervention comparing gratitude journaling and neutral journaling on subjective happiness and stress levels.

# Data Overview
```{r laoding data{}
setwd("C:/Users/0&1/OneDrive/Documents/Student-Projects-Portfolio/gratitude_wellbeing")
data <- read.csv("gratitude_wellbeing_data.csv")
head(data)
summary(data)
```

- Groups: Gratitude vs. Neutral

- Measures:

   - Happiness: Pre and Post

   - Stress: Pre and Post

   - Compliance over 14 days
   
# Mixed-Effects Model: Happiness Over Time
We fit a mixed-effects model with random intercepts per subject.

```{r}
data_long <- data %>%
  pivot_longer(cols = c(Happiness_Pre, Happiness_Post),
               names_to = "Time", values_to = "Happiness") %>%
  mutate(Time = factor(Time, levels = c("Happiness_Pre", "Happiness_Post")),
         SubjectID = factor(SubjectID))

model <- lmer(Happiness ~ Time * Group + (1 | SubjectID), data = data_long)
summary(model)
```

```{r}
library(lmerTest)
summary(lmer(Happiness ~ Group * Time + (1 | SubjectID), data = data_long))
```

# Result
A significant Time × Group interaction was found (p = 0.0011), indicating steeper gains in happiness for the gratitude group.

# Independent t-Test: Stress Reduction
```{r}
data <- data %>%
  mutate(Stress_Reduction = Stress_Pre - Stress_Post)

t_test <- t.test(Stress_Reduction ~ Group, data = data)
t_test
```

# Result
Stress levels decreased more in the Gratitude group than the Neutral group (p = 0.019).


Before the intervention, both groups had similar happiness scores — but by the end, the Gratitude group jumped from an average of 4.4 to 5.7, while the Neutral group only rose to 5.0. 

```{r}
library(ggplot2)

ggplot(data_long, aes(x = Time, y = Happiness, color = Group, group = Group)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  labs(title = "Change in Happiness Over Time by Group") +
  theme_minimal()
```



# Comparison of Compliance 

```{r}
# Compare compliance between groups
t.test(Compliance_14Day ~ Group, data = data_long)
```

```{r}
library(ggplot2)

ggplot(data_long, aes(x = Group, y = Compliance_14Day, fill = Group)) +
  geom_boxplot() +
  labs(title = "Compliance by Group", y = "Days Journaled (out of 14)") +
  theme_minimal()
```


# Handling Missing Data (Example)
Let's assume some missing happiness scores and use multiple imputation.

```{r}
set.seed(123)
data_miss <- data
data_miss$Happiness_Post[sample(1:nrow(data), 5)] <- NA

imp <- mice(data_miss, m = 5, method = "pmm", seed = 500)
summary(imp)

# Pool model
completed_data <- complete(imp, 1)
head(completed_data)
```

#Assumption Checks

```{r}
par(mfrow=c(1,2))
plot(model)
qqnorm(resid(model)); qqline(resid(model))
```

# References
- Emmons, R. A., & McCullough, M. E. (2003). Counting blessings versus burdens.

- Lyubomirsky, S., Sheldon, K. M., & Schkade, D. (2005). Pursuing happiness.

- Field, A. (2013). Discovering Statistics Using R.

# Conclusion
The gratitude journaling group experienced significantly greater increases in happiness and reduced stress levels over the 2-week period. These results support the use of gratitude interventions to enhance psychological well-being.
