---
title: "Perfectionism and Procrastination Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(psych)
library(lavaan)
library(semPlot)
library(interactions)
```

## Load Data

```{r load-data}
data <- read.csv("perfectionism_procrastination_data.csv")
kable(head(data))
```

## Descriptive Statistics

```{r descriptives}
describe(data)
```

## Path Analysis / SEM

```{r sem-model}
model <- '
  # regression paths
  Procrastination ~ SelfOriented + SociallyPrescribed + OtherOriented
'
fit <- sem(model, data = data)
summary(fit, fit.measures = TRUE, standardized = TRUE)
```

## SEM Path Diagram

```{r plot-sem}
semPaths(fit, whatLabels = "std", layout = "tree", edge.label.cex = 1.2, fade = FALSE)
```

## Moderation Analysis: Self-Oriented × GPA

```{r moderation}
# Center variables
mod_data <- data %>%
  mutate(SelfCentered = scale(SelfOriented, scale = FALSE),
         GPACentered = scale(GPA, scale = FALSE),
         Interaction = SelfCentered * GPACentered)

# Fit model with interaction
mod_model <- lm(Procrastination ~ SelfCentered * GPACentered + SociallyPrescribed + OtherOriented, data = mod_data)
summary(mod_model)

# Visualize interaction (if significant)
interact_plot(mod_model, pred = SelfCentered, modx = GPACentered, plot.points = TRUE)
```

## Interpretation

- Self-Oriented Perfectionism should significantly predict procrastination.
- Socially Prescribed is expected to be non-significant.
- GPA should not significantly moderate the relationship.
- Check CFI and RMSEA to determine model fit.

Let me know if you'd like this exported as a `.Rmd` file or zipped with the dataset!
